\ourmethodplaintext: Visualizing Function Type Signatures in R

### 2020-07-08 paper/main.tex
Update paper

%\documentclass{vgtc}                          % final (conference style)
\documentclass[review]{vgtc}                 % review
%\documentclass[widereview]{vgtc}             % wide-spaced review
%\documentclass[preprint]{vgtc}               % preprint
%\documentclass[electronic]{vgtc}             % electronic version

@@

@@

\usepackage{caption}
\captionsetup[table]{skip=10pt}
\usepackage{tabularx}
\usepackage{todonotes} %TODO: Remove after all TODO's are gone
\usepackage{lineno}
\linenumbers

%% ===================================== Commands
\newcommand{\code}[1]{\texttt{\small{#1}}}
\newcommand{\typevis}{{\sc TypeVis}\xspace}
\newcommand{\figref}[1]{Figure~\ref{#1}\xspace}
\newcommand{\pkgNumAnalyzed}{412\xspace}
\newcommand{\settingsTab}{{\sc Settings}\xspace}

@@

\title{Type Visualization for R}

\author{Cameron Moy\thanks{camoy@ccs.neu.edu}
\and Julia Belyakova\thanks{belyakovay@ccs.neu.edu}}
\affiliation{\scriptsize Northeastern University}

\abstract{
  Data-driven approaches to programming language design are uncommon.
  Despite the availability of large code repositories, distilling useful
  information from programs remains difficult. Important dimensions,
  like run-time type data, are inscrutable without the appropriate tools.
  \typevis is an interactive visualization for exploring and analyzing
  type information from program execution traces---aiding
  user understanding of function type signatures across many executions.
  The system uses flows to intuitively represent categorical
  and numerical aspects of type signatures.
  Insights derived from our visualization are aimed at informing
  language design decisions---specifically of a new gradual type system
  currently being developed for the R programming language.
}

@@

Programming languages commonly evolve by decree. Often, the language designer
decides that a new feature is necessary, or that a past feature was
ill-conceived. Thus, the language moves forward---
forcing its users to adapt to the changes.
Sometimes, these changes are the result of rigorous theory
or the product of community feedback. However, rarely is language design
informed by empirical data on how programs are \emph{actually
written} in practice. Why so?

%For a popular language, data collection is no challenge. Vast
Thanks to the prevalence of open source code, collecting data on
the usage of a popular programming language is feasible. Vast
quantities of publicly available code are hosted on repositories such as Github.
Furthermore, if there is concern about the quality of this data,
such as code duplication~\cite{lopes:2017},
code can be obtained from curated
language-specific package repositories like npm or CRAN.

To inform programming language design,
the collected data needs to be analyzed and interpreted.
Programs are complex and highly structured,
so simple semantically-ignorant metrics like ``lines of code'' carry little significance.
Therefore, researchers often employ static and dynamic analyses
to gather information about specific aspects of programs.
Even then, it may be difficult to make sense of the
results of these analyses,
especially if the data set is large.
%Rather, the difficulty lies with analyzing and interpreting the data
%at hand. Code is complex and highly structured. Simple

To analyze just one aspect of programming languages, \emph{type usage}, we
built \typevis, an interactive visualization of run-time type signatures.
%an interactive visualization of run-time types for the language R.
\typevis was created with the R programming language in mind,
and builds on a data set of execution traces
mined from the unit tests and vignettes
of the most widely used libraries in the R ecosystem.

The tool supports three broad tasks:
\begin{itemize}
  \setlength\itemsep{0em}
  \item filtering the data set down to an amenable subset;
  \item understanding the types of function arguments and returns;
  \item comparing type signatures across different functions.
\end{itemize}
%For filtering, \typevis uses the treemap idiom
%to convey the relative scale of different packages.
For exploring and filtering the data set,
\typevis relies on treemaps to represent R packages and functions,
with node size conveying relative popularity.
Additionally, one can search directly for a function by name.
Once a subset of one or more functions is selected,
their type signatures are encoded as flows over types,
where flow width is proportional to frequency.
These flows can be readily compared within the same function,
or between different functions.

\typevis can assist language designers during multiple phases of development.
For example, exploratory data analysis can identify unexpected edge cases,
or weed out language designs that are incompatible with existing programs.
\typevis was built in support of a new gradual type system for R,
whose adoption depends on integrating well with existing code.
However, the visualization is not limited to R and can be used to
analyze type usage in any language where similar data are available.

Supplemental material, including the data and source code of
\typevis, is freely available: \todo{Add archival link}

\bgroup
\def\arraystretch{1.75}
\begin{table*}
  \centering
  \begin{tabularx}{\linewidth}{c|c|c|X}
    Domain Goals & Search Task & Query Task & \multicolumn{1}{c}{Abstract Task Description} \\
    \hline
    Find Function & Locate & Identify & Finding a function is a location task, where the target function is known, but its location within the package hierarchy is not. Once the desired function is found, a user must be able to identify data of interest. \\

    Determine Types & Browse & Identify & Determining types is a browse task, where the target type signature is unknown, but the location of the function of interest is known. While browsing, a user must be able to identify the frequencies of particular type signatures.\\

    Compare Signatures & Lookup & Compare & Comparing type signatures is a lookup task where both target type signatures have already been located. A user must be able to discern how often a signature is used compared to another, both within a function and across different functions.\\
  \end{tabularx}
  \caption{Domain goals and their mid-level and low-level task abstractions.}
  \label{tab:tasks}
\end{table*}
\egroup

@@

R is a lazy, multi-paradigm, dynamically-typed programming language
widely used for statistical computing~\cite{morandat:2012}.
Unlike statically-typed languages,
R code contains no type annotations
and never passes through a typechecker.
At run-time,
functions may be called with values of arbitrary type,
and may return values of arbitrary type.\footnote{
  We use the word ``type'' to mean what others may call a ``type tag.''
  This is a run-time concept---we do not assume a static type system
  because there is none. Indeed, the goal is to help design one!
  A ``type signature'' in this context is one particular configuration
  of argument and return types a function is called with at run-time.
}

@@

For statically-typed languages, the type system is usually
an inseparable part of the language. Since they are designed
at the same time, the type system precedes any code written
in the language.
Gradual type systems, by contrast, are often designed atop
an existing dynamically-typed language~\cite{tobin-hochstadt:2016}.
Many R programs already exist,
so a successful gradual type system must
work well with extant code.

We are not aware of any previous work that applies
visualization techniques for the purpose of programming language design.
However, there are existing publications on understanding
software through visualization.
Bohnet et al.~\cite{bohnet:2009} visualize massive program execution traces
to improve developer comprehension of large software systems.
Their tool applies pruning and summarization techniques to reduce trace size.
They focus on call topology,
describing the structure of function invocations,
rather than types.
Telea et al.~\cite{telea:2009} also develop a methodology for displaying call structure.
However, they visualize call graphs of systems in their entirety,
rather than individual execution traces.
\typevis has elements of both,
investigating the system as a whole,
but doing so via execution traces.
In addition to standard call graphs,
LaToza and Myers~\cite{latoza:2011} adorn nodes and edges with extra information,
allowing the user to gain further understanding of the system
under examination.
This includes grouping methods that come from the same class---a
basic form of type data.

@@

The data set we used comes from
Turcotte et al.~\cite{turcotte:2020}
and consists of traces from \pkgNumAnalyzed packages,
containing over $760,000$ lines of R code
and $534,000$ lines of native code.
All libraries were obtained from CRAN,
the most popular R code repository.
Packages hosted on CRAN are curated and satisfy
certain quality standards.
Out of the over $15,000$ packages
distributed on CRAN,
the subset of 412 packages considered here
all had code coverage above 65\%.
\todo{Confirm with Alexi this paragraph. Get proper BibTeX entry.}

Execution traces were recorded by running a dynamic analysis
over the test, example, and vignette code of each package.
A vignette is a form of documentation that weaves
together prose and executable code.
An execution trace is a list of function type signatures
containing, among other attributes, the function name
and type tags for each argument and return value.
For example, the function call {\tt !(FALSE)}
would be recorded as the type signature {\tt logical $\to$ logical},
meaning that the function {\tt !} took a logical value (i.e. a boolean)
and returned a logical value.
The type tags are assigned by the dynamic analysis
and refine the actual type tags at run-time.
For instance, the analysis
assigns the scalar tag {\tt double} to a singleton vector of double values
because R does not have native support for scalars.
The data set is a table of reduced execution traces.
Because \typevis was designed for the analysis of types,
not all information in the traces is relevant.
%that have been processed to reduce some fields.
%Traces were generated via a dynamic analysis
%that instruments the R virtual machine with hooks
%that intercept the required data at run-time.

While \typevis has only been instantiated with
data from R execution traces, the architecture is not limited to R.
Any dynamically-typed language, %with the right run-time
%hooks could generate data for the tool.
or statically-typed language with run-time tag information,
could be instrumented to generate data for the tool.
As long as the trace output adheres to
the correct schema, \typevis will work properly without any modifications.

@@

@@

We interviewed the primary researcher developing a gradual type system for R
to inventory relevant domain-specific tasks.
Additionally, we sought his and other researchers' feedback
throughout the creation of \typevis
to ensure it was adequately meeting requirements.
The domain-specific goals identified were:

@@

These goals can be classified using an abstract task analysis framework.
See table~\ref{tab:tasks} for a categorization of these domain tasks
with the mid-level and low-level abstract task taxonomy
established by Brehmer and Munzner~\cite{brehmer:2013}.

@@

\begin{figure*}
 \centering
 \includegraphics[width=\linewidth]{img/typevis.png}
 \caption{The main view of \typevis, with type flows on the left and the package hierarchy on the right. Here, the {\tt length.unit} and {\tt valid.unit} functions from the {\tt grid} package are both selected and displayed simultaneously.}
 \label{fig:typevis}
\end{figure*}

\typevis is an interactive web-based visualization
that permits real-time data exploration,
even with large data sets.
The default data set of R execution traces,
described in \S\ref{sec:data},
contains over a million rows.

\figref{fig:typevis} shows
the heart of \typevis---the type flow panel on the left
and the function filtering panel on the right.
Type signatures shown on the left reflect
the subset of the data selected on the right.
By default, if no subset is selected, the type flow
panel shows several of the most frequent function type signatures
in the entire data set.
%in its initial configuration.
Various aspects of the system,
such as the number of type signatures shown at once,
can be tweaked by the user if desired;
for this, a separate \settingsTab tab is provided.

\subsection{Function Filtering}

The right-hand panel of \figref{fig:typevis}
contains three elements: a search bar,
a package treemap,
and a function treemap.
All three components are linked.
Selecting a package
in the top treemap causes the bottom treemap
to show only functions defined in that package;
additionally, the type flow panel gets updated
with the most frequent function type signatures in the package.
Selecting a function in the bottom treemap
makes the type flow panel show only signatures
of that function.
Entering a function name in the search bar
updates the type flow panel immediately.

These different components support at least two modes
of search: quick function lookup and package exploration.
If a user has a specific function in mind,
they can type its name into the search bar and see its
type signatures straight away;
the search bar supports autocomplete that is
able to select one function among thousands.
If, on the other hand, one wants to browse
all available packages and functions,
the treemap view is more applicable.
The area of a block in the function treemap (package treemap)
conveys the log-scaled
frequency of calls to that function
(to all functions defined in that package).
%Area in the bottom treemap conveys the log-scaled
%frequency of function calls,
%and correspondingly the top treemap
%is scaled based on invocations
%of functions defined in that package.
Log-scaling and pagination keep
elements of the treemaps legible,
even when absolute size differences are significant.

The data set can additionally be filtered by analyzed package,
in other words, the package containing the call site of a function, not the definition site.
As mentioned in \S\ref{sec:data}, the data set contains information
from \pkgNumAnalyzed analyzed packages, which are all taken into account
by default. However, a user can limit the set of analyzed packages,
using a search bar in \settingsTab.

Although treemaps are often used for displaying
hierarchical data~\cite{shneiderman:1992},
in \typevis they have a flat structure.
The packages in our data are not deeply structured,
so the treemaps in \figref{fig:typevis} are only one level deep.
However, in many programming languages, like Java,
package systems contain deeply nested namespace hierarchies.
The treemap idiom is appropriate for navigation of such
hierarchies, and
%However, if the data set recorded deeply nested packages,
\typevis could be modified to handle that.

By default, \typevis allows only one function to be selected at a time,
but it can be configured for multiple function selection.
In the example from \figref{fig:typevis},
the {\tt length.unit} and {\tt valid.unit} functions
from the {\tt grid} package are both selected.
These functions are rendered simultaneously in the type flow
panel, allowing the user to make comparisons between them.
However, if the user wants to explore different functions rather than
compare them, the default one-function mode is more appropriate.

@@

Prominently featured in the center of \figref{fig:typevis}
is a parallel sets diagram encoding the type signatures of the selected
functions as flows.
Parallel sets,
proposed by Kosara et al.~\cite{kosara:2006},
are ideally suited for type signatures.
Each data point can be thought of as
a tuple $(f, \tau_1, \ldots, \tau_n, \tau_r)$
corresponding to the function name,
argument types,
and return type.
Every component of that tuple is categorical,
and we are interested in visualizing the frequencies of these points.
This is the exact use-case for parallel sets.

@@

In \figref{fig:typevis},
the {\tt length.unit} function is called with
two different type signatures at run-time.
Most frequently, the function is used with the
signature {\tt double $\to$ integer},
but a substantial fraction of the time it is
used as {\tt double[] $\to$ integer}.
Here, {\tt double[]} stands for an array of
doubles with arbitrary dimension.

Interactivity significantly enhances the
usability of the parallel sets diagram.
Hovering over a flow will highlight the
entire path of the flow
and will fade out all other functions,
as shown in \figref{fig:typevis}.
When displaying many flows at the same time,
highlighting becomes critical
for user comprehension.
Additional quantitative information
about the flow is supplied on-demand while hovering.
Clicking a flow will focus on that function
only, filtering away all others;
double clicking outside flows will bring back the previous view.
The number of flows shown at a time is limited,
and will paginate if that threshold is exceeded.

Flow layout is a necessary consideration
for a legible visualization~\cite{sugiyama:1981}.
Figure~\ref{fig:decross} exemplifies what
happens if flow layout is not adequately computed.
The top layout was generated without any
flow decrossing algorithm,
while the bottom layout applies optimal decrossing.
Even with highlighting,
the top visualization is almost impossible to understand
due to excessive flow overlap.
To maintain real-time performance,
we use two different decrossing methods.
First, the system attempts to construct
a layout with the globally optimal solution,
minimizing the amount of flow crossing by
solving a mixed-integer linear program~\cite{ebner:2005}.
Unfortunately, for even moderately
sized graphs this can take too long to compute.
Therefore, after about $1$ second,
the optimal algorithm is terminated
if it has not finished, and the final layout is computed
by a faster
algorithm that only minimizes local crossing between layers.
%computes the layout.
While the resulting layout might not be ideal,
it is much better than the one without decrossing at all.

@@

On the frontend, state management, interactivity,
and DOM manipulation is all done with the Vue framework.
This differs from many web-based visualizations
that use D3 for DOM manipulation.
We chose this setup because \typevis retains many distinct
components whose state must be kept synchronized.
This includes not only the visualizations, but also
user interface elements like the search bar.
In our experience, Vue's model of reactivity is
more successful at maintaining consistency across the
entire application than D3 and vanilla JavaScript.
While \typevis does not use D3 for DOM manipulation,
it still does use it for other visualization-related
tasks. Decrossing is computed using web workers
to allow early termination if the thread takes too long.

@@

@@

Designed with a specific use case in mind,
\typevis suffers from some shortcomings
and does not fully address the entire range of tasks
a language designer may need to perform.

In particular, \typevis does not allow for exploring 
type information at differing granularities.
R supports a range of data types
including ones containing dimensionality information.
For example, an R array may have type {\tt integer[3]}
indicating that it is an array of length $3$.
Every such type will get compressed down %into a single array of
to a simpler array
type {\tt integer[]}.
Reducing the number of types can make the visualization
more understandable, at the expensive of losing
precision in type information.
Ideally, a user would be able to interactively tune
the level of granularity based on their interest.
Such a feature would be especially helpful for R,
because values can come attached with lots of relevant
data such as the dynamic dispatch method
or user-specified metadata.

Filtering is a central mechanism of our system,
but it also comes at a cost.
If one has a specific function or several functions
in mind, then \typevis provides a useful
local view of that information.
It does not support, however, any kind of global
view of type information across a large number of
functions. When the amount of flows or nodes
becomes too great, the visualization will start
paginating, making comparisons across all flows impossible.
Aggregation and summarization would be key to
making a global view of the data feasible, but that remains future work.

>>>>

\documentclass{vgtc}                          % final (conference style)
%\documentclass[review]{vgtc}                 % review
%\documentclass[widereview]{vgtc}             % wide-spaced review
%\documentclass[preprint]{vgtc}               % preprint
%\documentclass[electronic]{vgtc}             % electronic version

@@

\usepackage[T1]{fontenc}% to use more chars & support copy & paste of special chars from PDF

@@

\usepackage{hyperref}
\usepackage[noabbrev]{cleveref}
\usepackage{caption}
\captionsetup[table]{skip=10pt}
\usepackage{tabularx}
\usepackage{todonotes} %TODO: Remove after all TODO's are gone

%% ORCID
\usepackage{scalerel}
\usepackage{tikz}
\usetikzlibrary{svg.path}

\definecolor{orcidlogocol}{HTML}{A6CE39}
\tikzset{
  orcidlogo/.pic={
    \fill[orcidlogocol] svg{M256,128c0,70.7-57.3,128-128,128C57.3,256,0,198.7,0,128C0,57.3,57.3,0,128,0C198.7,0,256,57.3,256,128z};
    \fill[white] svg{M86.3,186.2H70.9V79.1h15.4v48.4V186.2z}
                 svg{M108.9,79.1h41.6c39.6,0,57,28.3,57,53.6c0,27.5-21.5,53.6-56.8,53.6h-41.8V79.1z M124.3,172.4h24.5c34.9,0,42.9-26.5,42.9-39.7c0-21.5-13.7-39.7-43.7-39.7h-23.7V172.4z}
                 svg{M88.7,56.8c0,5.5-4.5,10.1-10.1,10.1c-5.6,0-10.1-4.6-10.1-10.1c0-5.6,4.5-10.1,10.1-10.1C84.2,46.7,88.7,51.3,88.7,56.8z};
  }
}

\newcommand\orcid[2]{\href{https://orcid.org/#2}{#1 \mbox{\scalerel*{
\begin{tikzpicture}[yscale=-1,transform shape]
\pic{orcidlogo};
\end{tikzpicture}
}{|}}}}

%% ===================================== Commands
\newcommand{\code}[1]{\texttt{\small{#1}}}
\newcommand{\ourmethodplaintext}{\textsc{TyVis}\xspace}
\newcommand{\ourmethod}{{\sc \ourmethodplaintext}\xspace}
\newcommand{\pkgNumAnalyzed}{412\xspace}
\newcommand{\settingsTab}{{\sc Settings}\xspace}

\newcommand{\osfsupplementplain}{osf.io/mc6zt}
\newcommand{\osfsupplement}{\formattedurl{https://\osfsupplementplain}{\osfsupplementplain}}
\newcommand{\osfsupplementlong}{\formattedurl{https://\osfsupplementplain}{https://\osfsupplementplain}}

\newcommand{\osfprereg}{\formattedurl{https://\osfpreregplain}{\osfpreregplain}}

\newcommand{\formattedurl}[2]{\href{#1}{\small\texttt{#2}}\xspace}

@@

\title{\ourmethodplaintext: Visualizing Function Type Signatures in R}

\newcommand{\mailtohref}[1]{\href{mailto:camoy@ccs.neu.edu;belyakovay@ccs.neu.edu;alexi@ccs.neu.edu;dibartolomeo.s@northeastern.edu;c.dunne@northeastern.edu}{\color{black}#1}}

\author{
    \orcid{Cameron Moy}{0000-0002-4384-6351}%
    \thanks{%
        E-mails: \mailtohref{%
        [~%
        {camoy} |
        {belyakovay} |
        {alexi}~]@ccs.neu.edu,
        [~%
        {dibartolomeo.s} |
        {c.dunne}~]@northeastern.edu}
        \vspace{0.55in}
    }\\%
    \and \orcid{Julia Belyakova}{0000-0002-7490-8500}\\%
    \and \orcid{Alexi Turcotte}{0000-0002-0381-0477}\\%
    \and \orcid{Sara Di Bartolomeo}{0000-0001-9517-3526}\\%
    \and \orcid{Cody Dunne}{0000-0002-1609-9776}\\%
}
\affiliation{\scriptsize Northeastern University}

\teaser{
  \includegraphics[width=\textwidth]{img/teaser.png}
  \caption{Type signatures for a subset of R's base package functions shown in the type flow visualization of \ourmethod.
  The function name is listed at the top followed by the first two argument types.
  All argument and return types are shown in the full visualization.}
}

\abstract{
  Data-driven approaches to programming language design are uncommon.
  Despite the availability of large code repositories, distilling semantically-rich information from programs remains difficult.
  Important dimensions, like run-time type data, are inscrutable without the appropriate tools.
  We contribute \ourmethod, an interactive visualization for exploring and analyzing type information from program execution traces --- aiding user understanding of function type signatures across many executions.
  The system uses flows to intuitively represent categorical and numerical aspects of type signatures.
  Insights derived from our visualization are aimed at informing language design decisions --- specifically of a new gradual type system currently being developed for the R programming language.
  A full version of this paper with all supplemental material is available at \osfsupplementlong
}

@@

\bgroup
\def\arraystretch{1.75}
\begin{table*}
  \centering
  \begin{tabularx}{\linewidth}{c|c|c|X}
    \textbf{Domain Goals} & \textbf{Search Task} & \textbf{Query Task} & \multicolumn{1}{c}{\textbf{Abstract Task Description}} \\
    \hline
    Find function      & Locate & Identify & 
    Finding a function is a locate task, where the target function is known, but its location within the package hierarchy is not.
    Once the desired function is found, a user must be able to identify relevant data of interest. \\

    Determine types    & Browse & Identify & 
    Determining types is a browse task, where the target type signature is unknown, but the location of the function of interest is known.
    While browsing, a user must be able to identify the frequencies of particular type signatures.\\

    Compare signatures & Lookup & Compare  & 
    Comparing type signatures is a lookup task where both target type signatures have already been located.
    A user must be able to discern how often a signature is used compared to another, both within a function and across different functions.\\
  \end{tabularx}
  \caption{
  Domain goals for programming language designers trying to understand function type signatures at run time.
  We captured these goals from interviews and created mid-level and low-level task abstractions based on Brehmer \& Munzner's multi-level task typology~\cite{brehmer:2013}.}
  \label{tab:tasks}
\end{table*}
\egroup

Programming languages commonly evolve by decree.
Often, the language designer decides that a new feature is necessary, or that a past feature was ill-conceived.
Thus, the language moves forward --- forcing its users to adapt to the changes.
Sometimes, these changes are the result of rigorous theory or the product of community feedback.
However, rarely is language design informed by empirical data on how programs are \emph{actually written} in practice~\cite{hanenberg:2010}.
Why so?

Thanks to the prevalence of open source code, collecting data on the usage of a popular programming language is feasible.
Vast quantities of code are publicly available online on language-specific package servers like npm for JavaScript or CRAN for R.
Some repositories, like CRAN, only include high-quality libraries.
Using a curated source of code mitigates some data collection concerns like code duplication~\cite{lopes:2017}.

To inform programming language design, the collected data needs to be analyzed and interpreted.
Programs are complex and highly structured, so simple semantically-unaware metrics like ``lines of code'' carry little significance.
Therefore, researchers often employ static and dynamic analyses to gather information about specific aspects of programs.
Even then, it may be difficult to make sense of the results of these analyses, especially if the data set is large.

Following the design study ``lite'' methodology~\cite{syeda:2020}, we built \ourmethod, an interactive visualization of run-time type signatures.
A type signature is a description of the argument and return types that a particular function is called with at run-time.
\ourmethod was created with the R programming language in mind, and builds on a data set of execution traces mined from the unit tests and vignettes of the most widely used libraries in the R ecosystem.

The tool supports three broad tasks:
\begin{itemize}
  \setlength\itemsep{0em}
  \item filtering the data set down to an amenable subset;
  \item understanding the types of function arguments and returns;
  \item comparing type signatures across different functions.
\end{itemize}

For exploring and filtering the data set, \ourmethod relies on treemaps to represent R packages and functions, with node size conveying relative popularity.
Additionally, one can search directly for a function by name.
Once a subset of one or more functions is selected, their type signatures are encoded as flows over types, where flow width is proportional to frequency.
These flows can be readily compared within the same function, or between different functions.

\ourmethod can assist language designers during multiple phases of development.
For example, exploratory data analysis can identify unexpected edge cases, or weed out language designs that are incompatible with existing programs.
\ourmethod was built in support of a new gradual type system for R, whose adoption depends on integrating well with existing code.
A type system provides a conservative approximation of run-time types --- thus, insight into existing type usage can inform decision-making.
Without a data-driven tool, gradual type system designers are left to guess about how their language is used in practice.
Our aim is to eliminate such guesswork.
While our visualization was created for a specific use case, it is not limited to R and can be used to analyze type usage in any language where similar data are available.
Even beyond this domain, linking parallel sets and treemaps can be a powerful visualization technique.
For example, parallel sets can display survey responses with many categorical questions~\cite{kosara:2006}, but a large number of responses can introduce too much visual clutter.
Hierarchical filtering, with a treemap, can eliminate irrelevant information and recover an interpretable representation.

@@

R is a lazy, multi-paradigm, dynamically-typed programming language widely used for statistical computing~\cite{morandat:2012}.
Unlike statically-typed languages, R code contains no type annotations and never passes through a typechecker.
At run-time, functions may be called with values of arbitrary type, and may return values of arbitrary type.\footnote{
  We use the word ``type'' to mean what others may call a ``type tag.''
  This is a run-time concept --- we do not assume a static type system   because there is none.
  Indeed, the goal is to help design one!
  A ``type signature'' in this context is one particular configuration of argument and return types a function is called with at run-time.
}

@@

For statically-typed languages, the type system is usually an inseparable part of the language.
Since they are designed at the same time, the type system precedes any code written in the language.
Gradual type systems, by contrast, are often designed atop an existing dynamically-typed language~\cite{tobin-hochstadt:2016}.
Many R programs already exist, so a successful gradual type system must work well with extant code.
Thus, gradual type system designers must first understand existing language idioms, and then work to accommodate them.
Our tool assists during the understanding phase of development.

We are not aware of any previous work that applies visualization techniques for the purpose of programming language design.
However, there are existing publications on understanding software through visualization.
Bohnet et al.~\cite{bohnet:2009} visualize massive program execution traces to improve developer comprehension of large software systems.
Their tool applies pruning and summarization techniques to reduce trace size.
They focus on call topology, describing the structure of function invocations, rather than types.
Telea et al.~\cite{telea:2009} also develop a methodology for displaying call structure.
However, they visualize call graphs of systems in their entirety, rather than individual execution traces.
\ourmethod has elements of both, investigating the system as a whole, but doing so via execution traces.
In addition to standard call graphs, LaToza and Myers~\cite{latoza:2011} adorn nodes and edges with extra information, allowing the user to gain further understanding of the system under examination.
This includes grouping methods that come from the same class --- a basic form of type data.

@@

We used the data set from Turcotte et al.~\cite{turcotte:2020} that consists of traces from over $760,000$ lines of R code and $534,000$ lines of native code.
This code was obtained from CRAN, a curated repository of R packages that is widely used in the community.
The data set consists of all packages with at least $65\%$ code coverage and at least $5$ reverse dependencies (clients using that package).
Out of the over $15,000$ packages distributed on CRAN, this amounted to \pkgNumAnalyzed packages.

Execution traces were recorded by running a dynamic analysis over the test, example, and vignette code of each package.
A vignette is a form of documentation that weaves together prose and executable code.
An execution trace is a list of function type signatures containing, among other attributes, the function name and type tags for each argument and return value.
For example, the expression {\tt 0 < 1} would be recorded as the type signature {\tt integer $\to$ integer $\to$ logical}, meaning that the function {\tt <} took two integer values and returned a logical value (i.e. a boolean).
The type tags are assigned by the dynamic analysis and refine the actual type tags at run-time.
For instance, the analysis assigns the scalar tag {\tt double} to a singleton vector of double values because R does not have native support for scalars.
The data set is a table of reduced execution traces.
Because \ourmethod was designed for the analysis of types, not all information in the traces is relevant.

While \ourmethod has only been instantiated with data from R execution traces, the architecture is not limited to R.
Any dynamically-typed language, or statically-typed language with run-time tag information, could be instrumented to generate data for the tool.
As long as the trace output adheres to the correct schema, \ourmethod will work properly without any modifications.

@@

\begin{figure*}
 \centering
 \includegraphics[width=\linewidth]{img/typevis.png}
 \caption{An overview of \ourmethod, with type flows on the left and the package hierarchy on the right. Here, the {\tt length.unit} and {\tt valid.unit} functions from the {\tt grid} package are both selected and displayed simultaneously.}
 \label{fig:typevis}
\end{figure*}

@@

We interviewed the primary researcher developing a gradual type system for R (and third author of this work) to inventory relevant domain-specific tasks.
Additionally, we sought his and other researchers' feedback throughout the creation of \ourmethod to ensure it was adequately meeting requirements.
The domain-specific goals identified were:

@@

These goals can be classified using an abstract task analysis framework.
See \cref{tab:tasks} for a categorization of these domain tasks with the mid-level and low-level abstract task taxonomy established by Brehmer and Munzner~\cite{brehmer:2013}.

@@

\ourmethod is an interactive web-based visualization that permits real-time data exploration, even with large data sets.
The default data set of R execution traces, described in \cref{sec:data}, contains over a million rows.

\Cref{fig:typevis} shows the heart of \ourmethod --- the type flow panel on the left and the function filtering panel on the right.
Type signatures shown on the left reflect the subset of the data selected on the right.
By default, if no subset is selected, the type flow panel shows several of the most frequent function type signatures in the entire data set.
Various aspects of the system, such as the number of type signatures shown at once, can be tweaked by the user if desired; for this, a separate \settingsTab tab is provided.

\subsection{Hierarchical Package Visualization}

The right-hand panel of \cref{fig:typevis} contains three elements: a search bar, a package treemap, and a function treemap.
All three components are linked.
Selecting a package in the top treemap causes the bottom treemap to show only functions defined in that package; additionally, the type flow panel gets updated with the most frequent function type signatures in the package.
Selecting a function in the bottom treemap makes the type flow panel show only signatures of that function.
Entering a function name in the search bar updates the type flow panel immediately.

These different components support at least two modes of search: quick function lookup and package exploration.
If a user has a specific function in mind, they can type its name into the search bar and see its type signatures straight away; the search bar supports autocomplete that is able to select one function among thousands.
If, on the other hand, one wants to browse all available packages and functions, the treemap view is more applicable.
The area of a block in the function treemap (package treemap) conveys the log-scaled frequency of calls to that function (to all functions defined in that package).
Log-scaling and pagination keep elements of the treemaps legible, even when absolute size differences are significant.

The data set can additionally be filtered by analyzed package, in other words, the package containing the call site of a function, not the definition site.
As mentioned in \cref{sec:data}, the data set contains information from \pkgNumAnalyzed analyzed packages, which are all taken into account by default.
However, a user can limit the set of analyzed packages, using a search bar in \settingsTab.

Although treemaps are often used for displaying hierarchical data~\cite{shneiderman:1992}, in \ourmethod they have a flat structure.
The packages in our data are not deeply structured, so the treemaps in \cref{fig:typevis} are only one level deep.
However, in many programming languages, like Java, package systems contain deeply nested namespace hierarchies.
The treemap idiom is appropriate for navigation of such hierarchies, and \ourmethod could be modified to handle that.

By default, \ourmethod allows only one function to be selected at a time, but it can be configured for multiple function selection.
In the example from \cref{fig:typevis}, the {\tt length.unit} and {\tt valid.unit} functions from the {\tt grid} package are both selected.
These functions are rendered simultaneously in the type flow panel, allowing the user to make comparisons between them.
However, if the user wants to explore different functions rather than compare them, the default one-function mode is more appropriate.

@@

Prominently featured in the center of \cref{fig:typevis} is a parallel sets diagram encoding the type signatures of the selected functions as flows.
Parallel sets, proposed by Kosara et al.~\cite{kosara:2006}, are ideally suited for type signatures.
Each data point can be thought of as a tuple $(f, \tau_1, \ldots, \tau_n, \tau_r)$ corresponding to the function name, argument types, and return type.
Every component of that tuple is categorical, and we are interested in visualizing the frequencies of these points.
This is the exact use-case for parallel sets.

@@

In \cref{fig:typevis}, the {\tt length.unit} function is called with two different type signatures at run-time.
Most frequently, the function is used with the signature {\tt double $\to$ integer}, but a substantial fraction of the time it is used as {\tt double[] $\to$ integer}.
Here, {\tt double[]} stands for an array of doubles with arbitrary dimension.

Interactivity significantly enhances the usability of the parallel sets diagram.
Hovering over a flow will highlight the entire path of the flow and will fade out all other functions, as shown in \cref{fig:typevis}.
When displaying many flows at the same time, highlighting becomes critical for user comprehension.
Additional quantitative information about the flow is supplied on-demand while hovering.
Clicking a flow will focus on that function only, filtering away all others; double clicking outside flows will bring back the previous view.
The number of flows shown at a time is limited, and will paginate if that threshold is exceeded.

Flow layout is a necessary consideration for a legible visualization~\cite{sugiyama:1981}.
\Cref{fig:decross} exemplifies what happens if flow layout is not adequately computed.
The top layout was generated without any flow decrossing algorithm, while the bottom layout applies optimal decrossing.
Even with highlighting, the top visualization is almost impossible to understand due to excessive flow overlap.
To maintain real-time performance, we use two different decrossing methods.
First, the system attempts to construct a layout with the globally optimal solution, minimizing the amount of flow crossing by solving a mixed-integer linear program~\cite{ebner:2005}.
Unfortunately, for even moderately sized graphs this can take too long to compute.
Therefore, after about $1$ second, the optimal algorithm is terminated if it has not finished, and the final layout is computed by a faster algorithm that only minimizes local crossing between layers.
While the resulting layout might not be ideal, it is much better than the one without decrossing at all.

@@

On the frontend, state management, interactivity, and DOM manipulation is all done with the Vue framework.
This differs from many web-based visualizations that use D3 for DOM manipulation.
We chose this setup because \ourmethod contains many distinct components whose state must be kept synchronized.
This includes not only the visualizations, but also user interface elements like the search bar.
Vue's model of reactivity can easily maintain consistent state across the entire application.
While \ourmethod does not use D3 to directly manipulate the DOM, it still does use it for other visualization-related computations.
Decrossing is calculated using web workers to allow early termination if the thread takes too long.

@@

\section{Evaluation}

In accordance with the design study ``lite'' methodology~\cite{syeda:2020}, we conducted a small-scale qualitative usability study and received feedback from $18$ individuals. Many respondents had some trouble interpreting the visualization or at least commented that the learning curve was steep. Given that the domain is niche and the flow-based idiom for visualizing signatures is novel, this was not unexpected.

To alleviate some of this confusion, we incorporated elements that give explicit explanations for different components. These appear as question mark icons that, upon hovering, give a complete description of the corresponding component. For more details, we also provide a full page of step-by-step instructions.

%% TODO: Alexi will possibly add evaluation notes here.

@@

Designed with a specific use case in mind, \ourmethod suffers from some shortcomings and does not fully address the entire range of tasks a language designer may need to perform.

In particular, \ourmethod does not allow for exploring type information at differing granularities.
R supports a range of data types including ones containing dimensionality information.
For example, an R array may have type {\tt integer[3]} indicating that it is an array of length $3$.
Every such type will get compressed down to a simpler array type {\tt integer[]}.
Reducing the number of types can make the visualization more understandable, at the expensive of losing precision in type information.
Ideally, a user would be able to interactively tune the level of granularity based on their interest.
Such a feature would be especially helpful for R, because values can come attached with lots of relevant data such as the dynamic dispatch method or user-specified metadata.

Filtering is a central mechanism of our system, but it also comes at a cost.
If one has a specific function or several functions in mind, then \ourmethod provides a useful local view of that information.
It does not support, however, any kind of global view of type information across a large number of functions.
When the amount of flows or nodes becomes too great, the visualization will start paginating, making comparisons across all flows impossible.
Aggregation and summarization would be key to making a global view of the data feasible, but that remains future work.
